 

### 2023-05-24 markdown/methods.Rmd
explained information theory in concept of mutual information applied to the dataset validating normalizatin methods

>>>>

#### Validation of the methods applied to the dataset
In the context of the validation of methods applied to a dataset, the mutual information calculation is used to assess the impact of computations on the dataset and the information it carries. Specifically, the mutual information is calculated within each sample type to determine if the computations performed affect the dependence between samples of the same type.
Mutual information, denoted as $I(X;Y)$, is a measure of the dependence or information shared between two variables, $X$ and $Y$, in this case, two samples of the same sample type. It is calculated using the equation $I(X;Y) = H(X,Y) - H(X|Y) - H(Y|X)$, where $H$ represents the entropy.
Entropy, as a concept in information theory, quantifies the level of disorder or uncertainty associated with a random event. Events that are highly likely to occur are considered less informative than events that are more unexpected or less likely. In a venn diagramm $I(X;Y)$ (the mutual information) is the intersection between the two conditional Entropys $H(X|Y)$ and $H(Y|X)$.
By comparing the mutual information of a particular sample type before performing any calculations with the mutual information calculated after the computations, it is possible to assess whether the applied methods have significantly altered the dependence or information content within the sample type. The goal is to ensure that the computations do not completely decrease the information contributed by each biological sample type to the experiment.

The R package "infotheo" is utilized to perform the mutual information calculations, which is a commonly used package for information-theoretic analysis in R. Mutual information is returned as natural unit of information (=nat), a measurement proportional to the to the Shannon entropy ($1nat = \frac{1}{ln2} shannons$). Before calculating the pairwise mutual information of the expression matrix, values were discretized into bins with the discretize function of above package. After that all pairwise mutual information within the expression set before and after the pipeline is obtained. The pairwise differences can be observed as a boxplot for a particular sample type indicating the change in dependence within this sample type.
